---
title: "STA302 Final Project Part 1"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{r, message=FALSE, warning=FALSE}
# Clean data here
library(tidyverse)
data <- read.csv("Most Streamed Spotify Songs 2024.csv")

# Remove rows with empty cells
data[data == ""] <- NA
clean_data <- data %>% drop_na(Track, Album.Name, Artist, Release.Date, ISRC, 
                               Spotify.Streams, Spotify.Playlist.Count,
                               Spotify.Popularity, YouTube.Views, YouTube.Likes,
                               TikTok.Posts,TikTok.Likes, 
                               TikTok.Views, YouTube.Playlist.Reach,
                               Explicit.Track)

# Convert strings with commas into clean numerical values
cols <- c("Spotify.Streams", "Spotify.Playlist.Count", "Spotify.Playlist.Reach",
          "YouTube.Views", "YouTube.Likes", "TikTok.Posts",
          "TikTok.Likes", "TikTok.Views")

clean_data[cols] <- lapply(clean_data[cols], function(x) {
  x <- gsub(",", "", x)
  as.numeric(x)
})

# Add release decade column according to proposal categories 
clean_data$Release.Date <- as.Date(clean_data$Release.Date, format = "%m/%d/%Y")
clean_data$Release.Year <- as.numeric(format(clean_data$Release.Date, "%Y"))
clean_data$Release.Decade <- cut(
  clean_data$Release.Year,
  breaks = c(1900, 2009, 2019, 2029),
  labels = c("Pre-2010s", "2010s", "2020s")
)
```
```{r table-setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(knitr)
library(kableExtra)
library(tibble)
```

## Data Cleaning

Rows with empty cells were removed. Strings were converted into numbers. Release decade column was added.

```{r generate-table}
# 1.
table_data <- tibble::tribble(
  ~Variable, ~Mean, ~`Std. Dev.`, ~Median, ~Min, ~`Pctl. 25`, ~`Pctl. 75`, ~Max,
  "Spotify Streams (Millions)", "593.29", "552.71", "419.01", "0.01", "189.58", "819.62", "4281.47",
  "YouTube Views (Millions)",   "499.71", "755.93", "239.67", "0.28", "97.89",  "622.78", "16322.76",
  "YouTube Likes (Millions)",   "3.8",    "5.1",    "2.14",   "0.01", "0.86",   "4.72",   "62.31",
  "TikTok Posts (Thousands)",   "819.31", "2017.14","186.48", "0",    "44.09",  "722.14", "35323.8",
  "TikTok Likes (Millions)",    "122.82", "609.8",  "33.43",  "0",    "9.29",   "105.63", "23474.22",
  "TikTok Views (Millions)",    "1235.8", "6507.4", "316.04", "0",    "89.44",  "1009.27","233232.31",
  "Release Decade: Pre-2010s",  "2%",     "",       "",       "",     "",       "",       "",
  "Release Decade: 2010s",      "38.7%",  "",       "",       "",     "",       "",       "",
  "Release Decade: 2020s",      "59.4%",  "",       "",       "",     "",       "",       "")

# 2. Build the base table without any pipes
base_table <- kableExtra::kbl(
  table_data, 
  booktabs = TRUE, 
  align = c("l", rep("r", 7)),
  caption = "Summary Statistics of Predictors and Response Variables with N = 2778"
)

# 3. Apply styles one by one (No line-ending pipe risks)
styled_table <- kableExtra::kable_classic(base_table, full_width = TRUE, html_font = "Cambria")
styled_table <- kableExtra::row_spec(styled_table, 0, bold = TRUE)
styled_table <- kableExtra::row_spec(styled_table, 6, extra_css = "border-bottom: 1px solid black;")

# 4. Print the final table
styled_table
```


## Residual Plots

Below are the residual plots evaluating preliminary multiple linear regression
model assumptions.

```{r residual-plots, echo=FALSE, fig.height=4, fig.width=9}
# 1. Fit correct model matching proposal text variables
fit <- lm(
  Spotify.Streams ~
    YouTube.Views  * Release.Decade +
    YouTube.Likes  * Release.Decade +
    TikTok.Posts   * Release.Decade +
    TikTok.Likes   * Release.Decade +
    TikTok.Views   * Release.Decade, 
  data = clean_data
)
summary(fit) # Get intercept/coefficients
# 2. Arrange the two plots horizontally side by side in 1 row 2 column grid
par(mfrow = c(1, 2))
# 3. Draw plots cleanly without messy captions
# Residual vs Fitted plot
plot(fit, 
     which = 1, 
     sub.caption = ""
     )
# QQ plot
plot(fit,
     which = 2, 
     sub.caption = ""
     )
```


## Box-Cox transformation
```{r}
library(MASS)

# 1. Run the boxcox function on your existing model
bc <- boxcox(fit)

# 2. Extract the optimal lambda value
optimal_lambda <- bc$x[which.max(bc$y)]
print(optimal_lambda)
```


## Model Selection
```{r}
library(car)

# Transformed model
newfit <- lm(
  Spotify.Streams^(1/3) ~
    log(YouTube.Views)  * Release.Decade +
    log(YouTube.Likes)  * Release.Decade +
    log(TikTok.Posts)   * Release.Decade +
    log(TikTok.Likes)   * Release.Decade +
    log(TikTok.Views)   * Release.Decade, 
  data = clean_data
)

summary(newfit) # Get intercept/coefficients

# Transformed model with no interaction
newfit_no_interaction <- lm(
  Spotify.Streams^(1/3) ~
    log(YouTube.Views) +
    log(YouTube.Likes) +
    log(TikTok.Posts)  +
    log(TikTok.Likes) +
    log(TikTok.Views) +
    Release.Decade, 
  data = clean_data
)

summary(newfit_no_interaction)

anova(newfit, newfit_no_interaction)

# Arrange the two plots horizontally side by side in 1 row 2 column grid
par(mfrow = c(1, 2))
# Draw plots cleanly without messy captions
# Residual vs Fitted plot
plot(newfit, 
     which = 1, 
     sub.caption = ""
     )
# QQ plot
plot(newfit,
     which = 2, 
     sub.caption = ""
     )
# Best subset selection
library(leaps)

best_subset <- regsubsets(Spotify.Streams ~ YouTube.Views + YouTube.Likes + TikTok.Posts + TikTok.Likes + TikTok.Views + Release.Decade, data = clean_data, nvmax = 10, nbest = 1, really.big=TRUE, method="exhaustive")

par(mfrow=c(1,2))

plot(best_subset,scale="adjr2");plot(best_subset, scale="bic")

# Reduced model from best subset selection
best_subset_fit <- lm(
  Spotify.Streams^(1/3) ~
    log(YouTube.Likes) +
    log(TikTok.Views) +
    log(TikTok.Likes) +
    Release.Decade, 
  data = clean_data
)

vif(best_subset_fit)

# Determine whether to drop TikTok views or TikTok Likes
fit_views <- lm(
  Spotify.Streams^(1/3) ~
    log(YouTube.Likes) +
    log(TikTok.Views) +
    Release.Decade, 
  data = clean_data_reduced
)

fit_likes <- lm(
  Spotify.Streams^(1/3) ~
    log(YouTube.Likes) +
    log(TikTok.Likes) +
    Release.Decade, 
  data = clean_data_reduced
)

AIC(fit_views, fit_likes)
BIC(fit_views, fit_likes)
summary(fit_views)$adj.r.squared
summary(fit_likes)$adj.r.squared
```
## Final Model Selection
```{r}
library(car)
finalfit <- lm(
  Spotify.Streams^(1/3) ~
    log(YouTube.Likes)  * Release.Decade +
    log(TikTok.Likes)   * Release.Decade +
    log(TikTok.Views)   * Release.Decade, 
  data = clean_data
)

summary(finalfit) # Get intercept/coefficients
# 2. Arrange the two plots horizontally side by side in 1 row 2 column grid
par(mfrow = c(1, 2))
# 3. Draw plots cleanly without messy captions
# Residual vs Fitted plot
plot(finalfit, 
     which = 1, 
     sub.caption = ""
     )
# QQ plot
plot(finalfit,
     which = 2, 
     sub.caption = ""
     )
```


```{r residual-plots-2, echo=FALSE, fig.height=6, fig.width=9, fig.align="center", fig.pos="H"}
residuals_fit <- residuals(newfit)
predictors <- list(
  "YouTube Views"  = clean_data$YouTube.Views,
  "YouTube Likes"  = clean_data$YouTube.Likes,
  "TikTok Posts"   = clean_data$TikTok.Posts,
  "TikTok Likes"   = clean_data$TikTok.Likes,
  "TikTok Views"   = clean_data$TikTok.Views
)

par(mfrow = c(2, 3))

for (name in names(predictors)) {
  x <- predictors[[name]]
  plot(
    x, residuals_fit,
    main  = paste("Residuals vs\n", name),
    xlab  = name,
    ylab  = "Residuals",
    pch   = 20,
    cex   = 0.5,
    col   = rgb(0, 0, 0, 0.3)
  )
  abline(h = 0, col = "gray", lwd = 1, lty = 2)
  lines(lowess(x, residuals_fit), col = "red", lwd = 1.5)
}

hist(
  residuals_fit,
  breaks = 50,
  main   = "Histogram of Residuals",
  xlab   = "Residuals",
  col    = "lightgray",
  border = "white"
)
abline(v = 0, col = "red", lwd = 2, lty = 2)
```

## Identifying Leverage Points, Outliers, and Influential Points
```{r}
n <- nrow(clean_data)
p <- length(coef(newfit)) - 1  # number of predictors (excluding intercept)

# Leverage Points
hat_values <- hatvalues(newfit)
lev_threshold <- 2 * (p + 1) / n
leverage_points <- which(hat_values > lev_threshold)
cat("Leverage threshold:", lev_threshold, "\n")
cat("Number of leverage points:", length(leverage_points), "\n")

# Outliers (Standarized Residuals)
stan_resid <- rstandard(newfit)
outlier_threshold <- 4
outliers <- which(abs(stan_resid) > outlier_threshold)
cat("\nOutlier threshold (|standardized residual| >", outlier_threshold, ")\n")
cat("Number of outliers:", length(outliers), "\n")

# Cook's Distance
cooks <- cooks.distance(newfit)
cooks_threshold <- 4 / n
influential_cooks <- which(cooks > cooks_threshold)
cat("\nCook's Distance threshold:", cooks_threshold, "\n")
cat("Number of influential points (Cook's D):", length(influential_cooks), "\n")

# DFFITS
dffits_vals <- dffits(newfit)
dffits_threshold <- 2 * sqrt((p + 1) / n)
influential_dffits <- which(abs(dffits_vals) > dffits_threshold)
cat("\nDFFITS threshold:", dffits_threshold, "\n")
cat("Number of influential points (DFFITS):", length(influential_dffits), "\n")

# DFBETAS
dfbetas_vals <- dfbetas(newfit)
dfbetas_threshold <- 2 / sqrt(n)
# Flag any observation that exceeds the threshold for ANY coefficient
influential_dfbetas <- which(apply(abs(dfbetas_vals) > dfbetas_threshold, 1, any))
cat("\nDFBETAS threshold:", dfbetas_threshold, "\n")
cat("Number of influential points (DFBETAS):", length(influential_dfbetas), "\n")

# Summary: Points flagged by all three influence measures
influential_all <- Reduce(intersect, list(influential_cooks, influential_dffits, influential_dfbetas))
cat("\nPoints flagged by Cook's D, DFFITS, and DFBETAS:", length(influential_all), "\n")

# Refit without most serious influential points
clean_data_reduced <- clean_data[-influential_all, ]

newfit_reduced <- lm(  
  Spotify.Streams^(1/3) ~
    log(YouTube.Views)  * Release.Decade +
    log(YouTube.Likes)  * Release.Decade +
    log(TikTok.Posts)   * Release.Decade +
    log(TikTok.Likes)   * Release.Decade +
    log(TikTok.Views)   * Release.Decade, 
  data = clean_data_reduced
)

# Compare refitted model and previous model (with influential points)
par(mfrow = c(2, 2))
plot(newfit,         which = 1, main = "Original",      sub.caption = "")
plot(newfit_reduced, which = 1, main = "Reduced",        sub.caption = "")
plot(newfit,         which = 2, main = "Original",      sub.caption = "")
plot(newfit_reduced, which = 2, main = "Reduced",        sub.caption = "")

AIC(newfit, newfit_reduced)
BIC(newfit, newfit_reduced)

# Adjusted R-squared and RSE side by side
cat("--- Original Model ---\n")
cat("Adj. R-squared:", summary(newfit)$adj.r.squared, "\n")
cat("RSE:", summary(newfit)$sigma, "\n")

cat("\n--- Reduced Model ---\n")
cat("Adj. R-squared:", summary(newfit_reduced)$adj.r.squared, "\n")
cat("RSE:", summary(newfit_reduced)$sigma, "\n")

# Extract removed rows
removed_points <- clean_data[influential_all, ]

# Inspect relevant columns for each removed point
removed_points[, c("Track", "Artist", "Release.Date", "Spotify.Streams",
                   "YouTube.Likes", "TikTok.Likes", "TikTok.Views", "Release.Decade")]
```
