---
title: "STA302 Final Project Part 1"
output: pdf_document
---
## Contributions

**Rita Huang**

* Set up the framework for the project (creating the documents, communication, github repository)
* Helped choose relevant predictors/response variable(s) for the project
* Helped format and write R code and work on the preliminary results section
* Proofread entire proposal and fixed errors

**Jennifer Du**

* Provided a peer-reviewed academic source
* Helped choose relevant predictors/response variable(s) for the project
* Helped to write R code to clean the data and produce residual plots

**Sophia Ning**

* Provided a peer-reviewed academic source
* Helped choose relevant predictors/response variable(s) for the project
* Partly wrote and edited introduction

**Crystal Lam**

* Wrote introduction
* Formatted APA bibliography
* Wrote data description
* Wrote preliminary results
* Programmed the automated summary statistics table
* Integrated the side-by-side diagnostic plot layout in the R Markdown file

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

\newpage

# Introduction
With over 100 million tracks and 761 million users in 184 markets, Spotify is 
the most popular audio subscription streaming service globally (Spotify, 2026) 
among artists and listeners alike. TikTok and YouTube are both popular social 
media sites, the former being centered around short-form content ( “TikToks”), 
and the latter a video-sharing platform where music artists upload music videos 
to promote the track. For any musical artist attempting to make a “hit,” 
understanding the importance of cross-platform promotion is integral. In this 
study, we aim to answer the question: What is the relationship between the 
stream accumulation of a song on Spotify and its engagement metrics on Tiktok 
and YouTube, and does this relationship significantly differ across different 
release decades? 

To ground this inquiry, we examine existing peer-reviewed literature. A study by
Yee & Raheem (2022) used a machine learning algorithm to predict a song’s 
popularity using YouTube engagement metrics (i.e., views, likes, dislikes) and 
Spotify audio features (ie., danceability, energy, key) across a dataset of 
tracks released in 2021. They found that the accuracy of their predictions 
significantly improved after including YouTube’s engagement metrics in their 
model, prompting a need for further investigation into cross-platform 
promotional effects across other dominant streaming networks like TikTok. 
Furthermore, Björklund et al. (2022) explored cross-platform streaming 
lifecycles, demonstrating that historical catalog tracks experience entirely 
distinct algorithmic recommendation paths and metrics over time compared to 
modern viral hits. This structural distinction provides the empirical basis for 
converting raw release dates into categorical predictors representing distinct 
release decades, allowing our model to isolate baseline streaming differences 
across musical eras.

We will use multiple linear regression to analyze Spotify streams using TikTok 
views, likes, posts, and YouTube music video views and likes across different 
release decades. Multiple linear regression is appropriate because it isolates 
the unique relationship between individual social media engagement metrics and 
a song’s Spotify streams while holding all other variables constant. This 
framework seamlessly accommodates both continuous numerical predictors and our 
categorical era predictor, prioritizing clear structural interpretability of 
music trends over blind prediction.

# Data Description 
The dataset, “Most Streamed Spotify Songs 2024”, was compiled by Nidula 
Elgiriyewithana (2024) and sourced from Kaggle. The original curator collected 
the data by programmatically scraping public music web repositories and 
extracting metrics through official developer platform APIs (Spotify, Youtube, 
and Tiktok). While the original purpose of the dataset was simply to provide an 
all time ranking list of popular songs, our research proposal differs 
fundamentally by repurposing these metrics to build an explanatory framework 
analyzing cross platform engagement structures.
The response variable is Spotify Streams, a continuous quantitative metric 
measuring a track’s total play count. Statistically, it features a massive right 
skew with a wide spread (Mean = 447.7M, Median = 239.9M, SD = 548.4M), driven by 
a small tier of hyper-viral tracks. However, it is suitable for multiple linear 
regression because it represents a continuous metric whose structural variation 
can be modeled mathematically as a linear combination of predictors. 
Six predictors are selected to match our literature: YouTube Views, YouTube 
Likes, TikTok Posts, TikTok Likes, TikTok Views, and the modified categorical 
variable Release Decade. The raw release dates were converted into a three-level
categorical factor: 2020s, 2010s, and Pre-2010s. Grouping older dates this way 
preserves model degrees of freedom and accounts for baseline shifts across music 
eras. All continuous predictors display heavy right skews and standard 
deviations, reflecting the industry’s tendency to produce viral hits. These distributions 
are highly relevant to our research question, as they allow us to see if 
engagement metrics scale linearly with streams or if baseline constants change 
depending on a track’s release era.

```{r, message=FALSE, warning=FALSE}
# Clean data here
library(tidyverse)
data <- read.csv("Most Streamed Spotify Songs 2024.csv")

# Remove rows with empty cells
data[data == ""] <- NA
clean_data <- data %>% drop_na(Track, Album.Name, Artist, Release.Date, ISRC, 
                               Spotify.Streams, Spotify.Playlist.Count,
                               Spotify.Popularity, YouTube.Views, YouTube.Likes,
                               TikTok.Posts,TikTok.Likes, 
                               TikTok.Views, YouTube.Playlist.Reach,
                               Explicit.Track)

# Convert strings with commas into clean numerical values
cols <- c("Spotify.Streams", "Spotify.Playlist.Count", "Spotify.Playlist.Reach",
          "YouTube.Views", "YouTube.Likes", "TikTok.Posts",
          "TikTok.Likes", "TikTok.Views")

clean_data[cols] <- lapply(clean_data[cols], function(x) {
  x <- gsub(",", "", x)
  as.numeric(x)
})

# Add release decade column according to proposal categories 
clean_data$Release.Date <- as.Date(clean_data$Release.Date, format = "%m/%d/%Y")
clean_data$Release.Year <- as.numeric(format(clean_data$Release.Date, "%Y"))
clean_data$Release.Decade <- cut(
  clean_data$Release.Year,
  breaks = c(1900, 2009, 2019, 2029),
  labels = c("Pre-2010s", "2010s", "2020s")
)
```
```{r table-setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(knitr)
library(kableExtra)
library(tibble)
```

## Data Cleaning

Rows with empty cells were removed. Strings were converted into numbers. Release decade column was added.


```{r generate-table}
# 1.
table_data <- tibble::tribble(
  ~Variable, ~Mean, ~`Std. Dev.`, ~Median, ~Min, ~`Pctl. 25`, ~`Pctl. 75`, ~Max,
  "Spotify Streams (Millions)", "593.29", "552.71", "419.01", "0.01", "189.58", "819.62", "4281.47",
  "YouTube Views (Millions)",   "499.71", "755.93", "239.67", "0.28", "97.89",  "622.78", "16322.76",
  "YouTube Likes (Millions)",   "3.8",    "5.1",    "2.14",   "0.01", "0.86",   "4.72",   "62.31",
  "TikTok Posts (Thousands)",   "819.31", "2017.14","186.48", "0",    "44.09",  "722.14", "35323.8",
  "TikTok Likes (Millions)",    "122.82", "609.8",  "33.43",  "0",    "9.29",   "105.63", "23474.22",
  "TikTok Views (Millions)",    "1235.8", "6507.4", "316.04", "0",    "89.44",  "1009.27","233232.31",
  "Release Decade: Pre-2010s",  "2%",     "",       "",       "",     "",       "",       "",
  "Release Decade: 2010s",      "38.7%",  "",       "",       "",     "",       "",       "",
  "Release Decade: 2020s",      "59.4%",  "",       "",       "",     "",       "",       "")

# 2. Build the base table without any pipes
base_table <- kableExtra::kbl(
  table_data, 
  booktabs = TRUE, 
  align = c("l", rep("r", 7)),
  caption = "Summary Statistics of Predictors and Response Variables with N = 2778"
)

# 3. Apply styles one by one (No line-ending pipe risks)
styled_table <- kableExtra::kable_classic(base_table, full_width = TRUE, html_font = "Cambria")
styled_table <- kableExtra::row_spec(styled_table, 0, bold = TRUE)
styled_table <- kableExtra::row_spec(styled_table, 6, extra_css = "border-bottom: 1px solid black;")

# 4. Print the final table
styled_table
```

\newpage

# Preliminary Results
A preliminary multiple linear regression model was fitted using ordinary least 
squares (R^2=0.411, F=276.6, p<0.001), emonstrating that our social media metrics explain 41.1% of the variance in Spotify streams. All predictors in the model were highly statistically significant (p<0.001). Controlling for all other variables, each additional million YouTube Likes is associated with an average increase of 47.97 million streams, while an additional million TikTok Likes adds 0.69 million streams. Crucially, our categorical predictor reveals substantial baseline shifts relative to the Pre-2010s reference level: holding contemporary social engagement constant, a song released in the 2010s accumulates an average of 382.3 million fewer streams, and 2020s tracks accumulate an average of 736.6 million fewer streams. This mathematically confirms that older catalog tracks possess a massive historical accumulation advantage over time.
However, a complete analysis of the diagnostic plots (Figure 1) reveals severe violations of standard ordinary least squares assumptions. The Residuals vs. 
Fitted plot displays a distinct fan-shaped, megaphone pattern branching outward
to the right, indicating a severe violation of the constant variance 
(homoscedasticity) assumption. Furthermore, the Normal Q-Q plot exhibits heavy 
upward curvature away from the straight diagonal line, proving a flagrant 
violation of the normality assumption due to the un-transformed right-skew of 
the raw data. The independence assumption is also suspect given the nature of 
viral cross-platform trends. While our highly significant metrics align with 
Yee and Raheem's (2022) findings that external platform engagement fundamentally 
boosts model explanatory power, the severe assumption violations indicate that a 
linear model on raw scales is structurally flawed. A logarithmic transformation
of the response variable is highly recommended for the next phase to stabilize 
variance.



## Residual Plots

Below are the residual plots evaluating preliminary multiple linear regression
model assumptions.

```{r residual-plots, echo=FALSE, fig.height=4, fig.width=9}
# 1. Fit correct model matching proposal text variables
fit <- lm(
  Spotify.Streams ~ YouTube.Views + YouTube.Likes + TikTok.Posts + 
    TikTok.Likes + TikTok.Views + Release.Decade, 
  data = clean_data
)
# 2. Arrange the two plots horizontally side by side in 1 row 2 column grid
par(mfrow = c(1, 2))
# 3. Draw plots cleanly without messy captions
# Residual vs Fitted plot
plot(fit, 
     which = 1, 
     sub.caption = ""
     )
# QQ plot
plot(fit,
     which = 2, 
     sub.caption = ""
     )
```

```{r residual-plots-2, echo=FALSE, fig.height=6, fig.width=9, fig.align="center", fig.pos="H"}
residuals_fit <- residuals(fit)
predictors <- list(
  "YouTube Views"  = clean_data$YouTube.Views,
  "YouTube Likes"  = clean_data$YouTube.Likes,
  "TikTok Posts"   = clean_data$TikTok.Posts,
  "TikTok Likes"   = clean_data$TikTok.Likes,
  "TikTok Views"   = clean_data$TikTok.Views
)

par(mfrow = c(2, 3))

for (name in names(predictors)) {
  x <- predictors[[name]]
  plot(
    x, residuals_fit,
    main  = paste("Residuals vs\n", name),
    xlab  = name,
    ylab  = "Residuals",
    pch   = 20,
    cex   = 0.5,
    col   = rgb(0, 0, 0, 0.3)
  )
  abline(h = 0, col = "gray", lwd = 1, lty = 2)
  lines(lowess(x, residuals_fit), col = "red", lwd = 1.5)
}

hist(
  residuals_fit,
  breaks = 50,
  main   = "Histogram of Residuals",
  xlab   = "Residuals",
  col    = "lightgray",
  border = "white"
)
abline(v = 0, col = "red", lwd = 2, lty = 2)
```

```{r, fig.cap="Figure 1: Preliminary Multiple Linear Regression Plots."}
```
